\bfseries \boldmath \scititle

### 2025-08-27 pyScripts/current.tex
update

% LaTeX template for Science journal paper
% Based on science_template.tex with adaptations for research content

@@

% Author and institution list.
% Institution numbers etc. should be hard-coded, do *not* use the \footnote command.
\author{
    Sarah M. Urbut$^{1,2,3,4\ast}$,
    Yi Ding$^{5}$,
    Tetsushi Nakao$^{1,2,4}$,
    Xilin Jiang$^{6}$,\and
    Leslie Gaffney$^{4}$,
    Anika Misra$^{1,2,4}$,
    Whitney Hornsby$^{1,2,4}$,
    Jordan W. Smoller$^{3,4,7,8}$\and
    Alexander Gusev$^{5,3,4\dagger}$,
    Pradeep Natarajan$^{1,2,3,4\dagger}$,
    Giovanni Parmigiani$^{9,10\dagger}$\and
    % Institution list, in a slightly smaller font
    \small$^{1}$Division of Cardiology, Department of Medicine, Massachusetts General Hospital, Boston, MA 02114, USA\and
    \small$^{2}$Cardiovascular Research Center, Massachusetts General Hospital, Boston, MA 02114, USA\and
    \small$^{3}$Harvard Medical School, Boston, MA 02215, USA\and
    \small$^{4}$Broad Institute of Harvard and MIT, Cambridge, MA 02142, USA\and
    \small$^{5}$Division of Population Sciences, Dana-Farber Cancer Institute, Boston, MA 02215, USA\and
    \small$^{6}$University of Cambridge, Cambridge, UK\and
    \small$^{7}$Psychiatric and Neurodevelopmental Genetics Unit, Center for Genomic Medicine, Massachusetts General Hospital, Boston, MA 02114,USA\and
    \small$^{8}$Department of Epidemiology, Harvard TH Chan School of Public Health, Boston, MA 02115, USA\and
    \small$^{9}$Department of Biostatistics, Harvard TH Chan School of Public Health, Boston, MA 02115, USA\and
    \small$^{10}$Department of Data Science, Dana-Farber Cancer Institute, Boston, MA 02215, USA\and    
    % Identify at least one corresponding author, with contact email address
    \small$^\ast$Corresponding author. Email: surbut@mgh.harvard.edu\and
    % Joint contributions can be indicated like this
    \small$^\dagger$These authors jointly supervised this work.
}
%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%

@@

\newpage
\begin{abstract} \bfseries \boldmath
Understanding how disease patterns evolve over a lifetime remains a key challenge in medicine. While electronic health records provide rich longitudinal data, existing models typically analyze each disease in isolation, missing the complex interplay between multiple conditions and genetic factors.

Here, we present \aladyn{}, a latent dynamic Bayesian framework that identifies latent disease signatures from longitudinal health records while modeling individual-specific trajectories. Applied to over 400,000 individuals across three biobanks with up to 50 years of follow-up, our model discovers clinically interpretable disease signatures that demonstrate strong consistency across diverse populations. These signatures reveal distinct progression patterns and show strong genetic correlations, enabling both accurate prediction of patient risk and discovery of novel genetic associations. This unified modeling approach significantly improves predictive performance, including an increase from 65\% to 90.1\% (median dynamic 10-year ASCVD prediction) in the receiver operator curve for coronary artery disease compared to traditional risk models, while simultaneously improving prediction for multiple diseases.

By integrating genetics with time-varying disease patterns, $\aladyn$ provides a unified approach to understanding disease comorbidity, identifies biological heterogeneity within clinical diagnoses, and offers new avenues for precision medicine through its ability to dynamically update risk assessments as new clinical information becomes available.
\end{abstract}

\section*{Main Text}

@@

Traditional approaches to analyzing EHR data often focus on isolated diseases or simple pairwise associations, failing to capture how multiple conditions evolve together over time \cite{wang_empirical_2021}. Recent unsupervised methods have attempted to identify disease clusters or trajectories \cite{jiang_age-dependent_2023}, but typically do not account for temporal dynamics of disease risk or individual-level heterogeneity, particularly the influence of genetic factors on disease progression rates \cite{urbut_dynamic_2023, hyttinen_genetic_2003}. Furthermore, many models assume conditional independence of diseases, missing the opportunity to leverage information across related conditions for both prediction and discovery \cite{blei_dynamic_2006, blei_latent_2003}.

We present $\aladyn$, a generative model that integrates genetic data with longitudinal EHRs to identify latent disease signatures while modeling individual-specific health trajectories over time. Our approach offers several key advantages over existing methods: (1) \textbf{Interpretability}: disease signatures correspond to clinically meaningful biological processes rather than abstract statistical factors; (2) \textbf{Temporal modeling}: captures how disease risk evolves dynamically over the life course rather than static risk assessment; (3) \textbf{Genetic integration}: directly incorporates genetic information into the model architecture rather than as post-hoc analysis; (4) \textbf{Unified framework}: simultaneously models multiple diseases, sharing information across related conditions and improving prediction even for diseases with sparse data \cite{caruana_multitask_1997}; and (5) \textbf{Individual-specific trajectories}: provides personalized risk profiles that adapt as new clinical information becomes available. By jointly modeling multiple diseases and their genetic determinants, $\aladyn$ enables both improved prediction of future disease risk and enhanced discovery of genetic architecture underlying complex phenotypes, while revealing meaningful patient subgroups with distinct biological mechanisms that could inform personalized interventions.

% Research Articles split the text into sections using descriptive headings

\subsection*{Model framework captures temporal disease signatures and individual trajectories}

Disease patterns among individuals vary by onset, progression speed, and composition, reflecting different underlying biological mechanisms. In contrast to topic models that conditionally allocate observed diseases to categories \cite{jiang_age-dependent_2023, blei_dynamic_2006}, $\aladyn$ models the probability of each disease for an individual by integrating across multiple latent signatures (Figure~\ref{fig:model_overview}).

For each individual $i$, disease $d$, and time point $t$, we model the probability of disease occurrence $\pi_{idt}$ as a mixture of $K$ signatures, each capturing patterns of diseases that tend to occur together,~so:

@@

\noindent where $\text{sigmoid}(x) = 1/(1+e^{-x})$, $\kappa$ is a global calibration parameter, $\theta_{ikt}$ represents a normalized individual $i$'s time-varying association with signature $k$ at time $t$, and $\phi_{kdt}$ captures the relationship between signature $k$ and disease $d$ over time.

@@

These latent variables $\lambda_{ikt}$ follow a Gaussian process \cite{Rasmussen2006Gaussian} prior wherein we model the effects of genetic factors and time. Specifically:

@@

\noindent where $r_k$ is a signature-specific reference level, $\Gamma_k$ captures genetic effects on signature predisposition, $\mathbf{g}_i$ represents individual genetic factors affecting the mean of $\lambda_{ik}$, and $\Omega_{\lambda}$ is a temporal covariance kernel modeling smooth trajectories for $\lambda_{ikt}$ over time.

@@

A key innovation of our approach lies in its formulation as a mixture of probabilities rather than a probability of a mixture, as in traditional sparse factor analysis approaches \cite{wang_empirical_2021}. Unlike allocation-based topic models that conditionally assign diseases to individuals $\textit{after}$ the event has necessarily occurred \cite{jiang_age-dependent_2023}, $\aladyn$ directly models the probability of disease occurrence as a weighted combination of signature-specific probabilities.

@@

\textbf{Two complementary applications:} $\aladyn$ serves two distinct but complementary purposes, each requiring different analytical approaches. For biomedical discovery, $\aladyn$ transforms understanding disease patterns, progression speed, genetic relationships, and disease associations—we leverage complete patient trajectories to maximize our ability to identify biological patterns and mechanisms. This retrospective analysis enables us to characterize disease signatures, quantify genetic influences, and reveal patient heterogeneity within diagnostic categories. For clinical prediction, we forecast future disease risk in real-world clinical settings—we employ a rigorous temporal validation framework that uses only information available up to a prediction time point (see \ref{fig:recruitment_timeline}). This prospective approach simulates real-world clinical scenarios where physicians must predict future risk based solely on a patient's history to date, ensuring our performance metrics reflect true predictive capability rather than retrospective explanation.

% Next heading
\subsection*{Applying $\aladyn$ identifies consistent signature patterns across diverse populations}

We applied $\aladyn$ to three independent cohorts: UK Biobank (UKB, n=427,239), Mass General Brigham (MGB, n=48,069), and All of Us (AoU, n=208,263). Despite differences in population characteristics, healthcare systems, and data collection methodologies, our model identified remarkably consistent signature patterns across these cohorts (Supplementary Table S1 for comparison). We obtained ICD-10 codes from set of hospitalization diagnoses in each biobank \cite{sudlow_uk_2015, koyama_decoding_2023} and transformed to respective pheCodes \cite{bastarache_using_2021}, following established approaches for EHR phenotyping \cite{hripcsak_next-generation_2013}. A set of 348 PheCodes were chosen representing diseases with at least 1000 unique occurrences in a list of 1.3 million UK Biobank hospitalization episode statistics \cite{yeung_ukbpheno_2022} as in \cite{jiang_age-dependent_2023}.

We set K=20 for our model, which then learned to identify 20 distinct disease signatures from the data. These model-derived signatures corresponded to recognized disease processes and captured diverse disease domains including cardiovascular, metabolic, pulmonary, psychiatric, musculoskeletal, and oncologic conditions. Each signature demonstrates characteristic temporal patterns, with disease probabilities evolving dynamically with age (Figure~\ref{fig:signature-dynamics}A). For example, the cardiovascular signature shows steadily increasing probabilities for conditions like atrial fibrillation and heart failure after age 55 years, while the malignancy signature displays a sharp rise in metastatic disease probabilities between ages 60-75 years. The specificity of each signature for a given disease, as modeled by $\psi_{kd}$, is preserved but heterogeneous, reflecting the model's ability to disentangle signature-disease specificity (Figure~\ref{fig:signature-dynamics}B).

Furthermore, the model's tensor structure (Figure~\ref{fig:data-description}), enables rapid disease hazard calculation using the average loadings (Equation~\ref{eq:lambda_gp}) and population-level $\phi_{kd}$. The average age-specific hazard probabilities for a wide range of diseases are visualized in Figure~\ref{fig:signature-dynamics}C, highlighting temporal risk patterns.

These signature patterns show strong consistency across the three independent cohorts (Figure~\ref{fig:signature-dynamics}D). When comparing the membership of diseases within signatures between any two cohorts, we observed high concordance (median modified Jaccard index = 0.792, IQR = 0.65-0.89 across all pairwise comparisons between biobanks for similarity-matched signatures when computing intersection among signatures normalized to total number of diseases within the matched UKB signature). Figure~\ref{fig:signature-dynamics}E illustrates this consistency for two key signatures: cardiovascular disease and malignancy. Despite differences in healthcare systems and coding practices, the temporal patterns of key diseases within these signatures remain remarkably consistent, supporting the biological validity of the discovered patterns.

The model also captures disease-specific temporal dynamics that match clinical expectations. For instance, Type 1 diabetes peaks earlier in life compared to Type 2 diabetes within the metabolic signature, while primary malignancies precede metastatic disease within the cancer signature. These nuanced temporal relationships emerge directly from the model without explicit encoding, demonstrating $\aladyn$'s ability to learn clinically meaningful disease trajectories.

\subsection*{Personalized trajectories reveal heterogeneity within disease categories}
Beyond population-level signatures, $\aladyn$ provides individual-specific trajectory information through the time-varying $\lambda_{ikt}$ parameters. These parameters capture each person's evolving association with different disease signatures throughout their life course. Figure~\ref{fig:individual_trajectories}A--C illustrate this for three representative patients (9493, 383060, and 148746) in UKBB, each of whom experienced myocardial infarction (MI). For each patient, the left panel shows how normalized signature loadings ($\theta$) dynamically change as new diagnoses are acquired, the middle panel displays the disease timeline, and the right panel summarizes the patient's average signature composition across all time points. The initially dominant signatures and the timing and magnitude of cardiovascular signature activation differ substantially between these individuals: for example, Patient 9493 shows a gradual increase in cardiovascular signature loading around age 60, Patient 148746 exhibits cardiovascular activation beginning subtly around age 50 and accelerating dramatically before MI at age 54, while Patient 383060 develops MI at age 72 after a more gradual cardiovascular signature evolution dominated by respiratory and dermatologic conditions. %These diverse patterns highlight how signature loadings are not static properties but dynamic quantities that respond to new clinical information, and how retrospective averages can obscure important differences in disease progression.

These examples demonstrate several key principles of the $\aladyn$ model: First, signature loadings are dynamic and often change in advance of formal diagnosis—note, for instance, the gradual rise in cardiovascular signature before clinical CAD diagnosis. Second, the model captures the interdependence between different disease processes, showing how the relative contribution of musculoskeletal, cardiovascular, and other signatures shift in relation to each other over the patient's life course. When comparing these individuals, we observe that while retrospective averages of signature loadings may appear similar, the plots of peak signature loading and the detailed trajectory shapes reveal substantial differences in disease progression patterns—differences that would be missed by models focusing only on time-averaged data.

Aggregating these individual patterns reveals distinct group-level differences. In a retrospective analysis, the comparison of early-onset ($\leq$ 55 years, mean age of event 49.7 years) and late-onset ($\geq$ 70 years, mean age of onset 74.9 years) MI in Figure~\ref{fig:individual_trajectories}D shows that early-onset patients exhibit a higher and earlier peak in cardiovascular signature contribution, as well as a more rapid increase in signature loading prior to the event, compared to late-onset cases. These quantitative differences in trajectory characteristics suggest that early- and late-onset MI, while sharing the same clinical diagnosis, may represent distinct disease entities requiring different preventive strategies.

Our model also illustrates how individual-level trajectories and population phenomena combine to elicit time-varying personalized disease probabilities. Figure~\ref{fig:individual_trajectories}E is a heatmap of log disease probabilities by signature and age for MI, showing how overall MI risk is decomposed into the contributions of various time-varying signature loadings. This visualization reveals the complex interplay between multiple signatures in determining disease risk. While the cardiovascular signature contributes most significantly to MI risk, other signatures—particularly those related to metabolic conditions and inflammation—also play important roles. The stacked area plot at the bottom shows how these contributions integrate to form the aggregate risk profile, revealing periods of accelerated risk accumulation that may represent critical windows for preventive intervention.

While each patient in Figures \ref{fig:individual_trajectories}A-C demonstrates MI, their disease journeys reveal biological differences among patients sharing this diagnosis, reflecting true heterogeneity—i.e., the presence of distinct subgroups with different underlying disease signature distributions—within the diagnostic category. Patient 148746 (Panel A) experiences MI at age 54 following a complex trajectory of gastrointestinal and musculoskeletal conditions, with cardiovascular signature activation beginning subtly around age 50 and accelerating dramatically in the years preceding the event. In contrast, Patient 383060 (Panel B) develops MI at age 72 after a markedly different prodrome dominated by respiratory and dermatologic conditions, showing a more gradual cardiovascular signature evolution. Post-MI trajectories also diverge substantially: Patient 148746 subsequently develops multiple cardiovascular complications and metabolic disorders, while Patient 383060's post-MI course is characterized by different comorbidity patterns including genitourinary and infectious disease manifestations. These distinct temporal signatures preceding and following identical clinical endpoints illustrate how $\aladyn$ captures the biological heterogeneity masked by traditional diagnostic categories—revealing that "myocardial infarction" encompasses diverse pathophysiological pathways that may require different prevention and treatment strategies.

This pattern of heterogeneity within diagnostic categories extends broadly across diseases. Figure \ref{fig:individual_trajectories}F captures signature heterogeneity within disease subtypes through stacked area plots showing deviations from the population average, highlighting the diversity of underlying biological processes among patients sharing the same clinical diagnosis.

To quantify this heterogeneity, we calculated cluster-specific standardized effect sizes, denoted $C_c$, by comparing, for each cluster $c$, the mean time-averaged signature loadings ($\theta$) among individuals assigned to cluster $c$ versus those assigned to all other clusters, using the Cohen's d formula\cite{cohen_statistical_2013} (see Figure~\ref{fig:cohen_d}). Here, $C_c$ quantifies the standardized difference in mean time-averaged signature loadings between patients in cluster $c$ and all other patients, providing a measure of how distinct each cluster is with respect to each disease signature. This analysis revealed substantial standardized differences in signature loadings between patient subgroups, reflecting biological processes not typically considered in diagnoses.

To systematically quantify the biological heterogeneity within each disease, we calculated Cohen's $d$ effect sizes and corresponding p-values for the difference in time-averaged signature loadings between clusters, across all signatures and for each of the three representative diseases (myocardial infarction, breast cancer, and major depressive disorder). This analysis revealed that the vast majority of signature differences between clusters were not only large in magnitude (with many Cohen's $d$ values exceeding 0.8, and some as high as 2.5–3.9), but also highly statistically significant (p $\leq 1\times$10$^{-8}$ for nearly all comparisons; see Supplementary Tables). For example, in major depressive disorder, the acute illness signature (Signature 16) showed a Cohen's $d$ of 3.87 (p $\approx$ 0) between clusters, indicating near-complete separation, while in myocardial infarction, the cardiovascular signature (Signature 5) reached a Cohen's $d$ of 2.96 (p $\approx$ 0). Similar patterns were observed for breast cancer, where the cardiovascular signature also showed strong differentiation (Cohen's $d$ = 2.46, p $\approx$ 0). Across all three diseases, most signatures exhibited highly significant differences between clusters, underscoring the presence of distinct biological subgroups within each diagnostic category. These results demonstrate that the observed heterogeneity is both quantitatively substantial and statistically robust, supporting the biological relevance of the identified patient subgroups.

>>>>

i% LaTeX template for Science journal paper
% Based on science_template.tex with adaptations for research content

@@

% Author and institution list.
% Institution numbers etc. should be hard-coded, do *not* use the \footnote command.
\author{
    Sarah M. Urbut$^{1,2,3,4}$,
    Yi Ding$^{5}$,
    Tetsushi Nakao$^{1,2,4}$,
    Xilin Jiang$^{6}$,\and
    Leslie Gaffney$^{4}$,
    Anika Misra$^{1,2,4}$,
    Whitney Hornsby$^{1,2,4}$,
    Jordan W. Smoller$^{3,4,7,8}$\and
    Alexander Gusev$^{5,3,4\dagger}$,
    Pradeep Natarajan$^{1,2,3,4\dagger}$,
    Giovanni Parmigiani$^{9,10\dagger}$\and
    % Institution list, in a slightly smaller font
    \small$^{1}$Division of Cardiology, Department of Medicine, Massachusetts General Hospital, Boston, MA 02114, USA\and
    \small$^{2}$Cardiovascular Research Center, Massachusetts General Hospital, Boston, MA 02114, USA\and
    \small$^{3}$Harvard Medical School, Boston, MA 02215, USA\and
    \small$^{4}$Broad Institute of Harvard and MIT, Cambridge, MA 02142, USA\and
    \small$^{5}$Division of Population Sciences, Dana-Farber Cancer Institute, Boston, MA 02215, USA\and
    \small$^{6}$University of Cambridge, Cambridge, UK\and
    \small$^{7}$Psychiatric and Neurodevelopmental Genetics Unit, Center for Genomic Medicine, \\ \small Massachusetts General Hospital, Boston, MA 02114,USA\and
    \small$^{8}$Department of Epidemiology, Harvard TH Chan School of Public Health, Boston, MA 02115, USA\and
    \small$^{9}$Department of Biostatistics, Harvard TH Chan School of Public Health, Boston, MA 02115, USA\and
    \small$^{10}$Department of Data Science, Dana-Farber Cancer Institute, Boston, MA 02215, USA\and    
    % Identify at least one corresponding author, with contact email address
    \small$^\ast$Corresponding author. Email: gp@jimmy.harvard.edu\and
    % Joint contributions can be indicated like this
    \small$^\dagger$These authors jointly supervised this work.
}
%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%

@@

\begin{abstract} \bfseries \boldmath
Understanding how disease patterns evolve over a lifetime remains a key challenge in medicine. While electronic health records provide rich longitudinal data, existing models typically analyze each disease in isolation, missing the complex interplay between multiple conditions and genetic factors. Here, we combine longitudinal health records with genetic data to model individual trajectories, using a novel latent dynamic Bayesian framework called $\aladyn$ that identifies latent disease signatures from longitudinal health records while modeling individual-specific trajectories. Applied across three biobanks with up to 50 years of follow-up, our model discovers clinically interpretable disease signatures that demonstrate remarkable consistency across diverse populations (79.2\% cross-cohort correspondence) and show strong genetic correlations, enabling both accurate prediction of patient risk and discovery of novel genetic associations. The model achieves dramatic improvements in disease prediction across 24 conditions (median dynamic AUC 0.85, gains up to 0.20), outperforming established clinical risk scores across diverse conditions. Furthermore, our signature-based approach identifies 150 novel genetic loci missed by single-disease GWAS, with multiple signatures showing strong genetic signals (cardiovascular h² = 0.041, musculoskeletal $h^{2}$ = 0.035). Critically, this unified modeling approach significantly improves predictive performance for multiple diseases while revealing distinct biological subtypes within traditional diagnostic categories—demonstrating substantial heterogeneity across diverse conditions including cancer, metabolic disorders, and psychiatric conditions, with Cohen's d effect sizes up to 3.87 for signature differences between patient clusters ($p \leq 1\times 10^{8}$ for 95\% of comparisons). In conclusion, $\aladyn$ combines genetics and longitudinal diagnosis to achieve both improved disease prediction and enhanced genetic discovery through a unified framework that captures the complex interplay between genetic predisposition and time-varying disease patterns.
\end{abstract}

\section*{Introduction}

@@

Traditional approaches to analyzing EHR data often focus on isolated diseases or simple pairwise associations, failing to capture how multiple conditions evolve together over time \cite{wang_empirical_2021}. Recent unsupervised methods have attempted to identify disease clusters or trajectories \cite{jiang_age-dependent_2023}, but typically do not account for temporal dynamics of disease risk or individual-level heterogeneity, particularly the influence of genetic factors on disease progression rates \cite{urbut_dynamic_2023, hyttinen_genetic_2003}. Furthermore, many models assume conditional independence of diseases, missing the opportunity to leverage information across related conditions for both prediction and discovery \cite{blei_dynamic_2006, blei_latent_2003}. Consider a patient who develops rheumatoid arthritis at age 45, followed by hypertension at 48, and eventually suffers a myocardial infarction at 52. Traditional approaches may treat these as separate events or simple comorbidities, missing the underlying metabolic-inflammatory signature that drives this progression. Also, they do not typically leverage information from patients with similar patterns to improve prediction for rare conditions, where limited data makes traditional disease-specific models less reliable.

We present $\aladyn$, a generative model that integrates genetic data with longitudinal EHRs to identify latent disease signatures while modeling individual-specific health trajectories over time.  $\aladyn$ addresses these limitations by identifying shared disease signatures that capture biological pathways common across multiple conditions, enabling more accurate prediction even for rare diseases through information sharing with related, more common conditions. Our approach offers several key advantages over existing methods: (1) \textbf{Interpretability}: disease signatures correspond to clinically meaningful biological processes rather than abstract statistical factors; (2) \textbf{Temporal modeling}: captures how disease risk evolves dynamically over the life course rather than static risk assessment; (3) \textbf{Genetic integration}: directly incorporates genetic information into the model architecture rather than as post-hoc analysis; (4) \textbf{Unified framework}: simultaneously models multiple diseases, sharing information across related conditions and improving prediction even for diseases with sparse data \cite{caruana_multitask_1997}; and (5) \textbf{Individual-specific trajectories}: provides personalized risk profiles that adapt as new clinical information becomes available. By jointly modeling multiple diseases and their genetic determinants, $\aladyn$ enables both improved prediction of future disease risk and enhanced discovery of genetic architecture underlying complex phenotypes, while revealing meaningful patient subgroups with distinct biological mechanisms that could inform personalized interventions.

% Research Articles split the text into sections using descriptive headings
\section*{Results}
\subsection*{$\aladyn$ captures temporal disease signatures and individual trajectories}

Disease patterns among individuals vary by onset, progression speed, and composition, reflecting different underlying biological mechanisms. Unlike allocation-based topic models that conditionally allocate observed diseases to categories \cite{jiang_age-dependent_2023, blei_dynamic_2006}, $\aladyn$  models the probability of each disease for an individual by integrating across multiple latent signatures (\textbf{Figure~\ref{fig:model_overview})}.

For each individual $i$, disease $d$, and time point $t$, we model the probability of disease occurrence $\pi_{idt}$ as a weighted combination of signature-specific probabilities, where each signature captures patterns of diseases that tend to occur together (\textbf{Table~\ref{tab:notation})}:

@@

\noindent where $\text{sigmoid}(\phi_{kdt}) = 1/(1+e^{-\phi_{kdt}})$, $\kappa$ is a global calibration parameter, $\theta_{ikt}$ represents a normalized individual $i$'s time-varying association with signature $k$ at time $t$, and $\phi_{kdt}$ captures the relationship between signature $k$ and disease $d$ over time $t$.

@@

These latent variables $\lambda_{ikt}$ follow a Gaussian process \cite{Rasmussen2006Gaussian} prior wherein we model the effects of genetic factors and time (see \textbf{Methods; Figure~\ref{fig:training_architecture})}. Specifically:

@@

\noindent where $r_k$ is a signature-specific reference level, $\Gamma_k$ captures genetic effects on signature predisposition, $\mathbf{g}_i$ represents individual genetic factors, here polygenic scores, affecting the mean of $\lambda_{ik}$, and $\Omega_{\lambda}$ is a temporal covariance kernel modeling smooth trajectories for $\lambda_{ikt}$ over time.

@@

A key innovation of our approach lies in its formulation as a mixture of probabilities rather than a probability of a mixture, as in traditional sparse factor analysis approaches \cite{wang_empirical_2021}. Unlike allocation-based topic models that conditionally assign diseases to individuals $\textit{after}$ the event has necessarily occurred \cite{jiang_age-dependent_2023}, $\aladyn$ directly models the probability of disease occurrence as a weighted combination of signature-specific disease probabilities.

@@

\textbf{Two complementary applications:} $\aladyn$ serves two distinct but complementary purposes, each requiring different analytical approaches. For biomedical discovery, $\aladyn$ operates with complete hindsight, leveraging entire patient trajectories to maximize our ability to identify biological patterns and mechanisms. This retrospective analysis transforms our understanding of disease patterns, progression speed, genetic relationships, and disease associations by using all available longitudinal data to characterize disease signatures, quantify genetic influences, and reveal patient heterogeneity within diagnostic categories. For clinical prediction, we operate under strict temporal constraints that mirror real-world clinical decision making (see \textbf{Figure~\ref{fig:training_architecture}} for distinction). We employ a rigorous temporal validation framework that uses only information available up to a prediction time point (see \textbf{Figure ~\ref{fig:recruitment_timeline}}). This prospective approach simulates real-world clinical scenarios where physicians must predict future risk based solely on a patient's history to date, ensuring our performance metrics reflect true predictive capability rather than retrospective explanation.

\subsection*{Applying $\aladyn$ identifies consistent signature patterns across diverse populations}

We applied $\aladyn$ to three independent cohorts: UK Biobank (UKB, n=427,239), Mass General Brigham (MGB, n=48,069), and All of Us (AoU, n=208,263) (\textbf{Table~ \ref{tab:baseline_characteristics}, Figure~\ref{fig:cohort_design}}). We obtained ICD-10 codes from hospitalization diagnoses in each biobank \cite{sudlow_uk_2015, koyama_decoding_2023} and transformed these to pheCodes \cite{bastarache_using_2021}, following established approaches for EHR phenotyping \cite{hripcsak_next-generation_2013} (see \textbf{Methods}). A set of 348 pheCodes were selected representing diseases with at least 1000 unique occurrences in UK Biobank hospitalization episode statistics \cite{yeung_ukbpheno_2022} as in \cite{jiang_age-dependent_2023}. Despite differences in population characteristics, healthcare systems, and data collection methodologies across these cohorts, our model identified remarkably consistent signature patterns (\textbf{Table\ref{tab:signature_summary_actual}; Figure~\ref{fig:cluster_correspondence})}.

We set K=20 for our model, which converged well (\textbf{\ref{fig:phi_batch}}) across all three biobanks and successfully identified 20 distinct disease signatures from the data. These model-derived signatures corresponded to recognized disease processes and captured diverse disease domains including cardiovascular, metabolic, pulmonary, psychiatric, musculoskeletal, and oncologic conditions (\textbf{\ref{tab:signature_summary_actual}}). Each signature demonstrates characteristic temporal patterns, with disease probabilities evolving dynamically with age across biobanks (\textbf{Figure~\ref{fig:signature-dynamics}A; Figures~\ref{fig:temporal_ukb},~\ref{fig:temporal_aou},~\ref{fig:temporal_mgb})}. For example, the cardiovascular signature shows steadily increasing probabilities for conditions like atrial fibrillation and heart failure after age 55 years, while the malignancy signature displays a sharp rise in metastatic disease probabilities between ages 60-75 years. The specificity of each signature for a given disease, as modeled by $\psi_{kd}$, is preserved but heterogeneous, reflecting the model's ability to disentangle signature-disease specificity (\textbf{Figure~\ref{fig:signature-dynamics}B)}.

Furthermore, the model's tensor structure (\textbf{Figure~\ref{fig:data-description})}, enables rapid disease hazard calculation using the average loadings (\textbf{Equation~\ref{eq:lambda_gp}}) and population-level $\phi_{kd}$. The average age-specific hazard probabilities for a wide range of diseases are visualized in \textbf{Figure~\ref{fig:signature-dynamics}C}, highlighting temporal risk patterns.

As stated, these signature patterns show strong consistency across the three independent cohorts (\textbf{Figure~\ref{fig:signature-dynamics}D,  Figures~\ref{fig:temporal_ukb},~\ref{fig:temporal_aou},~\ref{fig:temporal_mgb})}. When comparing the membership of diseases within signatures between any two cohorts, we observed high concordance (median modified Jaccard index = 0.792, IQR = 0.65-0.89 across all pairwise comparisons between biobanks for similarity-matched signatures when computing intersection among signatures normalized to total number of diseases within the matched UKB signature, \textbf{Figure ~\ref{fig:cluster_correspondence})}. \textbf{Figure~\ref{fig:signature-dynamics}E} illustrates this consistency for two key signatures: cardiovascular disease and malignancy. Despite differences in healthcare systems and coding practices, the temporal patterns of key diseases within these signatures remain remarkably consistent, supporting the biological validity of the discovered patterns.

The model also captures disease-specific temporal dynamics that match clinical expectations. For instance, Type 1 diabetes peaks earlier in life compared to Type 2 diabetes within the metabolic signature (\textbf{Fig \ref{fig:signature-dynamics}E}), while primary malignancies precede metastatic disease within the cancer signature. These nuanced temporal relationships emerge directly from the model without explicit encoding, demonstrating $\aladyn$'s ability to learn clinically meaningful disease trajectories.

\subsection*{Personalized trajectories reveal heterogeneity within disease categories}
Beyond population-level signatures, $\aladyn$ provides individual-specific trajectory information through the time-varying $\lambda_{ikt}$ parameters that reveal distinct disease progression patterns.

While each patient in \textbf{Figures \ref{fig:individual_trajectories}A-C} demonstrate similar average signature loadings in aggregate (horizonal 'static model summary' profile), their disease journeys reveal biological differences among patients sharing this diagnosis, reflecting true heterogeneity—i.e., the presence of distinct subgroups with different underlying disease signature distributions—within the diagnostic category. Patient C (Panel C) experiences MI at age 54 following a complex trajectory of gastrointestinal and musculoskeletal conditions, with cardiovascular signature activation beginning subtly around age 50 and accelerating dramatically in the years preceding the event. In contrast, Patient B (Panel B) develops MI at age 72 after a markedly different prodrome dominated by respiratory and dermatologic conditions, showing a more gradual cardiovascular signature evolution. Post-MI trajectories in these two patients also diverge substantially: Patient C subsequently develops multiple cardiovascular complications and metabolic disorders, while Patient B's post-MI course is characterized by different comorbidity patterns including genitourinary and infectious disease manifestations. These distinct temporal signatures preceding and following identical clinical endpoints illustrate how $\aladyn$ captures the biological heterogeneity masked by traditional diagnostic categories—revealing that "myocardial infarction" encompasses diverse pathophysiological pathways that may require different prevention and treatment strategies. Multiple additional examples (\textbf{Figure ~\ref{fig:patient_trajectories})} demonstrate the diversity in temporal loadings that would be missed by a summative approach considering only average loading.

Our model also illustrates how individual-level trajectories and population phenomena combine to elicit time-varying personalized disease probabilities. {\textbf{Figure~\ref{fig:individual_trajectories}E}} is a heatmap of log disease probabilities by signature and age for MI, showing how overall MI risk is decomposed into the contributions of various time-varying signature loadings. This visualization reveals the complex interplay between multiple signatures in determining disease risk. While the cardiovascular signature contributes most significantly to MI risk, other signatures—particularly those related to metabolic conditions and inflammation—also play important roles. The stacked area plot below demonstrates how these contributions integrate to form the aggregate risk profile, revealing periods of accelerated risk accumulation that may represent critical windows for preventive intervention.

Aggregating these individual patterns reveals distinct group-level differences. In a retrospective analysis, the comparison of early-onset ($\leq$ 55 years, mean age of event 49.7 years) and late-onset ($\geq$ 70 years, mean age of onset 74.9 years) MI in \textbf{Figure~\ref{fig:individual_trajectories}D} shows that early-onset patients exhibit a higher and earlier peak in cardiovascular signature contribution, as well as a more rapid increase in signature loading prior to the event, compared to late-onset cases. These quantitative differences in trajectory characteristics suggest that early- and late-onset MI, while sharing the same clinical diagnosis, may represent distinct disease entities requiring different preventive strategies.

This pattern of heterogeneity within diagnostic categories extends broadly across diseases. \textbf{Figure~\ref{fig:individual_trajectories}F} captures signature heterogeneity within disease subtypes through stacked area plots showing deviations from the population average, highlighting the diversity of underlying biological processes among patients sharing the same clinical diagnosis.

To systematically quantify differences in signature composition among patients with the same clinical diagnosis for three representative diseases (myocardial infarction, breast cancer, and major depressive disorder), we applied k-means clustering to patients' time-averaged signature loadings for each disease (\textbf{Figure~\ref{fig:individual_trajectories}F)} 
%This approach identifies distinct subgroups of patients who share similar underlying disease signature profiles despite having the same clinical diagnosis. 
\newcommand{\Ccsig}{$C_{ck}^{\mbox{\small SIG}}$ }
We then calculated cluster-specific Cohen's effect sizes\cite{cohen_statistical_2013} \Ccsig as follows (\textbf{Figure \ref{fig:cohen_d}; Extended data Data S1-S3)}. For cluster $c$ and  signature $k$, \Ccsig is the standardized difference in mean time-averaged signature loadings between individuals in cluster $c$ and those in all other clusters (see \textbf{Figure~\ref{fig:cohen_d})}. This measures of how distinct each cluster is with regard to each disease signature. %This analysis revealed substantial standardized differences in signature loadings between patient subgroups, reflecting biological processes not typically considered in diagnoses.
